A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-07 latex/main.tex

To validate the accuracy of times for non-sample internal nodes inserted 
into the ARG
(inferred using tsdate), we compared the dates of first emergence of
Nextstrain clades against the Nextstrain SARS-CoV-2 tree (downloaded on 
2023-01-21). Figure \ref{fig:pango_vs_nextstrain_node_times} shows relatively
close agreement for most clades, with only three (20C, 20E and 21B)
disagreeing by more than 28 days.

Then, to gauge internal consistency of the time estimation procedure, we
selected a subset of sample nodes, matching the distribution of node times 
and average number of children of the non-sample internal nodes. These
selected sample nodes were then re-dated, and the inferred dates compared 
against their known, true sampling dates. The results in Figure 
\ref{fig:node_time_testing} demonstrate that although the inferred and true
times are generally very close, there is a clear bias with re-estimated times
being consistently older (by an average of 37 days). This is in line with 
the findings of other recent
work \cite{sanderson2021chronumental}, and is likely to be a consequence
of edges carrying more mutations than would be expected under a neutral
molecular clock prior, as well as inaccuracies in the recording of 
sample dates.

@@

The B.1.1.529 lineage (Omicron) is believed to have emerged in southern Africa, and
it is a saltational lineage with two highly successful sublineages, BA.1 and BA.2 \cite{Dor2025}.
BA.1 and BA.2 are genetically distinct from each other and
from their parental B.1.1.529 lineage \cite{RambautOmicron2021}.
In the subgraph shown in Figure~\ref{fig:subgraph_omicron},
BA.1 and BA.2 differ from the non-Omicron B.1.1 ancestor
(above the B.1.1.529-labeled MRCA of BA.1 and BA.2)
by more than ten mutations, including six deleted regions.
The ARG captures the split between BA.1 and BA.2,
which has been shown in previous phylogenetic analysis \cite{Dor2025}.
The 5-base deletion highlighted in green above the BA.2 root
is an artefact of our naive deletion remapping procedure,
which splits two 9-base deletions in BA.1 and BA.2 into three deletions
(one in green at 11288, one in blue at 11283, and one in orange 11292;
see Document~\ref{sec:supplement_dels}).
Also, there is uncertainty in the sequence alignments
leading to ambiguity in the start position of these deletions,
as noted in the Pango designation issue \cite{RambautOmicron2021}.

We selected a BA.1 seed sample that led to the fewest mutations.
This strategy resulted in five mutations highlighted
in blue/orange/green/red/purple
which the chosen seed does not have
but are present in its inferred descendants.
In retrospect, SRR17051902 should have been a better choice for a seed,
because it would have resulted in a topology
where those five mutations would be placed by parsimony
above the split between BA.1 and BA.2.

@@

@@

\begin{figure}
\centering
\includegraphics[width=0.85\linewidth]{subgraph-Omicron.pdf}
\caption{
Subgraph illustrating the saltational origin of
the major Omicron lineages BA.1 and BA.2.
Symbols are described in Figure~\ref{fig:subgraph_delta},
except that BA.1 samples are filled in light blue
(also used to outline BA.1 characteristic mutations),
BA.2 samples in light yellow
(also used to outline BA.2 characteristic mutations), and
B.1.1.529 mutations
which are neither BA.1 or BA.2 characteristic mutations are outlined in green.
These colours correspond to those used
in the Pango designation issue \cite{RambautOmicron2021}.
}
\label{fig:subgraph_omicron}
\end{figure}

>>>>

To validate the accuracy of times for non-sample internal nodes inserted 
into the ARG
(inferred using tsdate), we compared the dates of first emergence of
Nextstrain clades against the Nextstrain SARS-CoV-2 tree (downloaded on 
2023-01-21). Figure \ref{fig:pango_vs_nextstrain_node_times} shows relatively
close agreement for most clades, with only three 
disagreeing by more than 28 days: 20C (an early lineage from 2020), 20E (Pango B.1.177 also known as EU1, emerging mid-2020), and 21B (Pango B.1.617.1 also known as Kappa, emerging in late 2020 to early 2021).

Then, to gauge internal consistency of the time estimation procedure, we
selected a subset of sample nodes, matching the distribution of node times 
and average number of children of the non-sample internal nodes. These
selected sample nodes were then re-dated, and the inferred dates compared 
against their recorded collection dates. The results in Figure 
\ref{fig:node_time_testing} demonstrate that although the inferred and true
times are generally very close, there is a clear bias with re-estimated times
being consistently older (by an average of 37 days). This is in line with 
the findings of other recent
work \cite{sanderson2021chronumental}, and is likely to be a consequence
of edges carrying more mutations than would be expected under a neutral
molecular clock prior (either due to errors in tree-building or genuinely varying mutation rates), as well as inaccuracies in the recording of 
sample dates.

@@

The B.1.1.529 lineage (Omicron)
is believed to have emerged in southern Africa, and
it is a saltational lineage with two highly successful sublineages,
BA.1 and BA.2 \cite{Dor2025}.
BA.1 and BA.2 are genetically distinct from each other and
from their parental B.1.1.529 lineage \cite{RambautOmicron2021}.
In the subgraph shown in Figure~\ref{fig:subgraph_omicron},
BA.1 and BA.2 differ from the non-Omicron B.1.1 ancestor
(above the B.1.1.529-labeled MRCA of BA.1 and BA.2)
by 27 mutations (including two deletions),
reflecting its saltational origin.
The ARG captures the split between BA.1 and BA.2
which has been shown in previous phylogenetic analysis \cite{Dor2025}.
The 5-base deletion highlighted in green above the BA.2 root
is an artefact of our naive deletion remapping procedure,
which splits two 9-base deletions in BA.1 and BA.2 into three deletions
(one in green at 11288, one in blue at 11283, and one in orange 11292;
see Document~\ref{sec:supplement_dels}).
Also, there is uncertainty in the sequence alignments
leading to ambiguity in the start position of these deletions
(if these deletions occur at 11283,
there would be a single deletion event before the BA.1/BA.2 split),
as noted in the Pango designation issue \cite{RambautOmicron2021}.

We selected a BA.1 seed sample that led to the fewest mutations.
This strategy resulted in five mutations
highlighted in blue/orange/green/red/purple
which are absent in the chosen
but present in its inferred descendants.
In retrospect, SRR17051902 should have been a better choice for a seed,
because it would have resulted in a topology
where those five mutations would be placed
above the split between BA.1 and BA.2.

@@

% Sc2ts detection results in matches
% https://github.com/jeromekelleher/sc2ts-paper/issues/366

% Multiple-breakpoint recombinants
% https://github.com/jeromekelleher/sc2ts-paper/issues/379

\begin{table}
\centering
\caption{
Sc2ts detection results for Pango X lineages not added to the ARG.
Some Pango X lineages present in the Viridian v04 dataset
were not integrated into the ARG absent in the ARG.
The number of samples in Viridian v04 dataset
assigned to a Pango X lineage and
the number of samples which passed sequence-level quality control
are shown here.
}
\label{tab:absent_pango_x}
\begin{tabular}{crrr}
\toprule
Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
\midrule
XBJ & 2 & 1 & 1\\
XBP & 8 & 2 & 1\\
XBS & 19 & 15 & 1\\
XBW & 1 & 1 & 1\\
XCA & 11 & 4 & 1\\
XAK & 2 & 2 & 1\\   % Multiple breakpoint Pango X
XAY & 15 & 11 & 6\\ % Multiple breakpoint Pango X
XBC & 36 & 21 & 3\\ % Multiple breakpoint Pango X
XBL & 2 & 2 & 0\\   % Multiple breakpoint Pango X
XBT & 1 & 1 & 2\\   % Multiple breakpoint Pango X
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=0.85\linewidth]{subgraph-Omicron_recolored.pdf}
\caption{
Subgraph illustrating the saltational origin of
the major Omicron lineages BA.1 and BA.2.
Symbols are described in Figure~\ref{fig:subgraph_delta},
except that BA.1 samples are filled in light blue
(also used to outline BA.1 characteristic mutations),
BA.2 samples in light yellow
(also used to outline BA.2 characteristic mutations), and
B.1.1.529 mutations
which are neither BA.1 or BA.2 characteristic mutations are
outlined in green.
These colors correspond to those used
in the Pango designation issue \cite{RambautOmicron2021}.
}
\label{fig:subgraph_omicron}
\end{figure}
